Predicting Personal Total Income of Individuals in the United States in 2023

### 2024-11-18 paper/paper.qmd
Revise Discussion and Render Paper

In this analysis, we explored three different models—Linear Regression, Random Forest, and XGBoost—to predict income based on various demographic and categorical factors. Each model provided valuable insights into the relationship between income and predictors such as education, age, industry, and gender. Among the models, XGBoost demonstrated the best performance in terms of predictive accuracy, as evidenced by its lower RMSE and superior handling of complex data interactions. However, this does not mean it is without limitations or that further improvements cannot be made.

@@

### Feature Importance
We have found that educational attainment and occupation industry are consistently the most important features in all three models, especially in XGBoost and Random Forest. This indicates that higher educational attainment and certain industries are correlated with higher personal income. This confirms our original hypothesis that certain demographic factors, such as education, may lead to specialized industry sectors where more income can be generated.

Age is also a useful predictor. However, despite the common belief that age is a determining factor when predicting personal income, its effectiveness is not as strong as education or industry. This may be caused by the fact that in some industries, your experience may hit a plateau, after which age may not contribute as much to experience as before certain age.

Marital status and sex do affect personal income, but not as significantly as the other features. It is great to prove that certain demographical features are not as informative when predicting personal income, suggesting a "fair" distribution of wealth.

@@

None of the three models used in our study provided an excellent prediction of personal income. For the residual analysis, it appears that the models' predictbility deterioate as the self-reported income gets higher. REsidual variance increased as the predicted income increased in the Random Forest and XGBoost model, suggesting potential heteroscedasticity. Our educated guess is that extra-high income may come from various sources, and simplifying them into a simple model like this is not a reasonable solution.

Among the three models, XGBoost has the highest accuracy overall. We believe that it can capture more complex relationships and interactions among the variables than the other two models. However, the high MSE for extra-high income individuals indicates that the model still needs to be refined more to achieve an accurate prediction.

Linear Regression has underperformed relative to the other two models. We speculate that simplicity in modeling linear relationships did not provide sufficient information to capture the non-linear patterns underlying our data, especially where multiple features interact to influence income, as suggested in @sec-afr.

@@

@@

-   While we focused on standard predictors like age, education, and industry, additional variables or interaction terms might improve model performance. For example, interactions between age and education, or industry and gender, could capture more nuanced relationships.

@@

-   The skewness in income distribution poses a significant challenge for modeling. Exploring techniques like synthetic data generation for high-income individuals or applying weighting schemes to account for data imbalance might help the models better capture the variance across all income levels.

In conclusion, while XGBoost provided the most accurate results, there is room for improvement in handling skewed data and better modeling high-income individuals. Future work will focus on addressing these issues while exploring more advanced modeling techniques and feature engineering to enhance predictive performance further.

>>>>

In this analysis, we explored three different models—Linear Regression, Random Forest, and XGBoost—to predict income based on various demographic and categorical factors. Each model provided valuable insights into the relationship between income and predictors such as education, age, industry, and gender. Among the models, XGBoost demonstrated the best performance in terms of predictive accuracy, as evidenced by its lower RMSE and superior handling of complex data interactions. However, this does not mean it is without limitations or that further improvements cannot be made. This paper aims to analyze the relationship between demographic and categorical factors and personal income through predictive modeling using Linear Regression, Random Forest, and XGBoost.

@@

We have found that educational attainment and occupation industry are consistently the most important features in all three models, especially in XGBoost and Random Forest. This indicates that higher educational attainment and certain industries are correlated with higher personal income. This confirms our original hypothesis that certain demographic factors, such as education, may lead to specialized industry sectors where more income can be generated.

Age is also a useful predictor. However, despite the common belief that age is a determining factor when predicting personal income, its effectiveness is not as strong as education or industry. This may be caused by the fact that in some industries, your experience may hit a plateau, after which age may not contribute as much to experience as before certain age. It would be interesting for future studies to explore non-linear relationships between age and income by incorporating interaction terms, such as age and education or age and industry.

Marital status and sex do affect personal income, but not as significantly as the other features. It is great to prove that certain demographical features are not as informative when predicting personal income, suggesting a "fair" distribution of wealth. However, the subtle influence of gender raises questions about income equity. While gender was less impactful overall, its specific effects in certain industries could be explored further.

@@

None of the three models used in our study provided an excellent prediction of personal income. For the residual analysis, it appears that the models' predictbility deterioate as the self-reported income gets higher. Residual variance increased as the predicted income increased in the Random Forest and XGBoost model, suggesting potential heteroscedasticity. Our educated guess is that extra-high income may come from various sources, and simplifying them into a simple model like this is not a reasonable solution.

Among the three models, XGBoost has the highest accuracy overall. We believe that it can capture more complex relationships and interactions among the variables than the other two models. However, the high MSE for extra-high income individuals indicates that the model still needs to be refined more to achieve an accurate prediction. Adding more predictors or considering advanced ensemble methods might help better account for the variability at higher income levels.

Linear Regression has underperformed relative to the other two models. We speculate that simplicity in modeling linear relationships did not provide sufficient information to capture the non-linear patterns underlying our data, especially where multiple features interact to influence income, as suggested in @sec-afr. It serves as a useful baseline but is less practical for this complex data structure.

@@

Additionally, the use of cross-sectional data in this analysis limits our ability to draw causal inferences. While the models identify associations, they do not account for temporal or dynamic effects, such as career progression, economic cycles, or policy changes, which might influence income over time. Longitudinal data could provide more robust insights into these dynamics.

@@

-   While we focused on standard predictors like age, education, and industry, additional variables or interaction terms might improve model performance. For example, interactions between age and education, or industry and gender, could capture more nuanced relationships. Adding geographic location or labor market data could also provide deeper insights into regional income disparities.

@@

-   Future studies could explore the temporal dimension of income predictors by incorporating longitudinal data, enabling an analysis of income trajectories over time. This would help assess not only predictors but also trends and changes in income inequality.

In conclusion, while XGBoost provided the most accurate results, there is room for improvement in handling skewed data and better modeling high-income individuals. Future work will focus on addressing these issues while exploring more advanced modeling techniques and feature engineering to enhance predictive performance further. A deeper dive into causal relationships and longitudinal data could also uncover additional factors influencing income dynamics.
